### 2013-06-11 text/Chapter04.tex
Refactor rules

\chapter{Structural coeffects} 
\label{ch:structural}

@@

@@

% ==================================================================================================
\newcommand{\ctxtrans}[2]{#1 \Rightarrow_c #2}
\begin{figure*}[t]
% Constants, if we wanted them, would look like this
%
% \inference[]
%   {c : \tau \in \Phi}
%   {\ctyp{1}{()} \vdash c : \tau}&
% \textnormal{\rulest{const}}
% \\[1.6em]

@@

\inference[]
  {}
  {\ctyp{0}{(v:\tau)} \vdash v : \tau }&
\textnormal{\rulest{var}}
\\[1.8em]
\inference
  {\ctyp{r \cprd s}{(\Gamma, v:\tau_1)} \vdash e : \tau_2 & v \notin \Gamma}
  {\ctyp{r}{\Gamma} \vdash \lambda v.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }&
\textnormal{\rulest{fun}}
\\[1.8em]
\inference
  {\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2 &
   \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1 }
  {\ctyp{s \cprd (r \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau_2}&
\textnormal{\rulest{app}}
\\[1.8em]
\inference
  {\ctyp{r \cprd s}{(\Gamma_1, v:\tau_1)} \vdash e_1 : \tau_2 &
  \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1}
  {\ctyp{r \cprd (s \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}&
\textnormal{\rulest{let}}
\\[1.8em]
\inference
  {\ctyp{r}{\Gamma} \vdash e : \tau &
   \ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}
  {\ctyp{r'}{\Gamma'} \vdash e : \tau}&
\textnormal{\rulest{ctx}}
\end{array}
\end{equation*}

@@

\begin{equation*}
\begin{array}{rcll}
\multicolumn{3}{c}{
  \inference
    {  \ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}} }
    {  \ctxtrans{\ctyp{r'\cprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)}} }
}&
\textnormal{\rulest{nest}}
\\[2.0em]
  \ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)} &}{& \ctyp{s\cprd r}{(\Gamma_2, \Gamma_1)}}&
\textnormal{\rulest{exch}}
\\[0.8em]
  \ctxtrans{ \ctyp{r \cprd 1}{(\Gamma, ())} &}{&
    \ctyp{r}{\Gamma}}&
\textnormal{\rulest{empty}}
\\[0.8em]
  \ctxtrans{\ctyp{r\cprd 0}{(\Gamma, x:\tau_1)} &}{& \ctyp{r}{\Gamma}}&
\textnormal{\rulest{weak}}
\\[0.8em]
  \ctxtrans{\ctyp{r \cvop s}{(x:\tau)} &}{&
    \ctyp{r \cprd s}{(x:\tau, x:\tau)}}&
\textnormal{\rulest{contr}}
\\[0.8em]
  \ctxtrans{\ctyp{r\cprd (s\cprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))} &}{&
    \ctyp{(r\cprd s)\cprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}\;&
\textnormal{\rulest{assoc}}
\\[0.8em]
\multicolumn{3}{c}{
  \ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}\quad(\textnormal{when}~s \cvop r = s)
}&
\textnormal{\rulest{sub}}
\end{array}
\end{equation*}

% We do not need this rule (can be expressed in terms of the other one)
%
% \inference
%  {  \ctxtrans{\ctyp{s}{\Gamma_2}}{\ctyp{s'}{\Gamma_2'}} }
%  {  \ctxtrans{\ctyp{r\cprd s}{(\Gamma_1, \Gamma_2)}}{\ctyp{r\cprd s'}{(\Gamma_1, \Gamma_2')}} }&
% \textnormal{\rulest{right}}

@@

As an example, consider a language that allows us to get a value of a variable (representing
some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
To track information about individual variables, we use a product-like operation $\cprd$ on tags 
to mirrors the product structure of variables. For example:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \cprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
  \vdash
    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
\end{array}
\end{equation*}
%
The coeffect tag $5 \cprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
a single variable:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \cvop 10}{(\ident{c}:\dnat)}
  \vdash
    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
\end{array}
\end{equation*}
%
In this example, the operation $\cvop$ would be the \emph{max} function and so $5 \cvop 10 = 10$.
Before looking at the formal definition, consider the typing of let bindings:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
\pastval{a}{15}+\pastval{c}{10}
\end{array}
\end{equation*}
%
The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.

The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
context is $\ctyp{0 \cprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
of $\cvop$ and it denotes empty coeffect. The typing context of the
body is $\ctyp{15 \cprd 10}{(\ident{a}, \ident{c})}$.

To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
of the context in which \ident{c} was defined using the $\cvop$ operation. This is then combined 
with the remaining tags from the body yielding the overall context:
$\ctyp{15 \cprd (10 \cvop (0 \cprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
normalization mechanism (described later), this can be further reduced to
$\ctyp{(15 \cvop 10) \cprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
\ident{b}.

@@

\subsection{Structural coeffect tags}
In the previous section, the $\cvop$ operation behaves similarly to the flat $\vee$ operation.
However, the type system does not require some of the semilattice properties, because some uses
are replaced with the $\cprd$ operation. We do not require any properties about the $\cprd$ operation.
For example, in the previous example cannot be commutative (since a tag $15 \cprd 10$ has different 
meaning than $10 \cprd 15$). However, we relate the operations using distributivity laws to allow 
normalization that was hinted above.

\begin{definition}
A \emph{structural coeffect tag structure} $(S, \cprd, \cvop, 0, 1)$ is a tuple where 
$(S, \cvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
by a binary operation $\cprd$ and element $1\in S$ such that
for all $r,s,t\in S$, the following equalities hold: 
%
\begin{subequations}
\begin{align}
\tag{associativity}
r \cvop (s \cvop t) &= (r \cvop s) \cvop t\\
\tag{commutativity}
r \cvop s &= s \cvop r\\
\tag{lower bound}
r \cvop 0 &= r\\
\tag{distributivity}
r \cvop (s \cprd t) &= (r \cvop s) \cprd (r \cvop t)\\
\tag{upper bound}
1 \cvop r &= 1
\end{align}
\end{subequations}
%
\end{definition}
%
% 'upper bound' (not sure what to call it) says that empty context cannot carry information
% looks like this is the upper bound of the bounded lattice ???
%
% commutativity is not strictly required by any law or type preservation, but
% it simplifies the presentation and looks reasonable
%
The tag $0$ represents that no coeffect is associated with a variable (i.e~when a variable is
always accessed using standard variable access). The tag $1$ is used to annotate empty 
variable context. For example, the context of an expression $\lambda x.x$ is empty and it needs
to carry an annotation. We explain exactly how this works when we introduce the type system.
The fact that $1$ is the upper bound means that combining it with other coeffect annotations
does not affect it and so empty contexts cannot carry any information.

@@

The simply typed system for \clstr~uses the same syntax of types as the system for \clflt. 
The rules are of a form $\ctyp{r}{\Gamma}\vdash e : \tau$ where $r$ is a tag provided by 
a \emph{structural coeffect tag structure} $(S,\cprd, \cvop, 0)$.

@@

As in linear and affine systems, the variable context $\Gamma$ in our system is not a 
simple set. Instead, we use the following tree-like structure (which is more similar to 
bunched types than to linear or affine systems):
%
\begin{equation*}
\Gamma ::= () \sep (x:\tau) \sep (\Gamma, \Gamma)
\end{equation*}
%
The syntax $()$ represents an empty context, so the structure defines a binary trees where
leaves are either variables or empty. Contexts such as $\ctyp{1\cprd(r \cprd 1)}{((), (x, ()))}$ 
contain unnecessary number of empty contexts $()$. However, we need to construct them
temporarily, because certain rules require splitting a context and, by our 
definition, the context $(x:\tau)$ is not splitable.

@@

The \rulest{app} rule combines coeffects $s, t, r$ associated with the function-returning
expression, argument and the function type respectively. 
The result of evaluating the argument in the context $t$ is passed to the function that
requires context $r$, so the variables used in the context $\Gamma_2$ are annotated with
the combination of coeffects $r \cvop t$. The variable context required to evaluate the 
function value is independent and so it is annotated just with coeffects $s$.
Finally, the \rulest{let} rule is derived from let binding and application, but we show
it separately to aid the understanding.

@@

\begin{lemma}[Definition of let binding]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau_2$
then $\ctyp{r}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$.
\end{lemma}
\begin{proof}
The premises and conclusions of a typing derivation of $(\lambda x.e_2)~e_1$
correspond with the typing rule \rulest{let}:
\begin{equation*}
\inference[]
  {\inference
    {\ctyp{r \cprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
    {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
   \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
  {\ctyp{r \cprd (s \cvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
\end{equation*}
\end{proof}

The term $e_2$ which is substituted in $e_1$ is checked in a different variable
and coeffect context $\ctyp{t}{\Gamma_2}$. This is common in sub-structural systems where
a variable cannot be freely used repeatedly. The context $\Gamma_2$ is used in place of 
the variable that we are substituting for. The let binding captures substitution for a 
specific variable (the context is of a form $\ctyp{r \cprd s}{\Gamma, v:\tau}$).
For a general substitution, we need to define the notion of context with a hole.

@@

\begin{equation*}
\begin{array}{rcl}
\Delta[-] &::=& \ctyp{1}{()} \\
       & \sep & \ctyp{r}{(x:\tau)}\\
       & \sep & \ctyp{-}{(-)} \\
       & \sep & \ctyp{r_1 \cprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
\end{array}
\end{equation*}
%
Assuming we have a context with hole $\ctyp{r}{\Gamma} \in \Delta[-]$, the hole filling 
operation $\chole{r}{\Gamma}{r'}{\Gamma'}$ fills the hole in the variable context with 
$\Gamma'$ and the corresponding coeffect tag hole with $r'$. The operation is defined in 
Figure~\ref{fig:structural-hole-fill}. Using contexts with holes, we can now formulate the
general substitution lemma for \clstr.

@@

\begin{equation*}
\begin{array}{rcl}
                              \chole{1}{()}{r'}{\Gamma'} &=& \ctyp{1}{()}\\
                        \chole{r}{(x:\tau)}{r'}{\Gamma'} &=& \ctyp{r}{(x:\tau)}\\
                             \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
\chole{r_1 \cprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \cprd r_2'}{(\Gamma_1', \Gamma_2')}\\
&\multicolumn{2}{l} {
  \textnormal{where}~~\ctyp{r_i'}{\Gamma_i'} = \chole{r_i}{\Gamma_i}{r'}{\Gamma'}
}
\end{array}
\end{equation*}
\caption{The definition of hole filling operation for $\Delta[-]$}
\label{fig:structural-hole-fill}
\vspace{-2em}
\end{figure}

\begin{lemma}[Substitution Lemma]
\label{thm:structural-subst}
If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
then $\chole{r}{\Gamma}{R \cvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
\end{lemma} 
\begin{proof}
Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
$(S, \cvop, 0, \cprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
\end{proof}

@@

\paragraph{Local soundness and completeness.}
As with the previous calculus, we want to guarantee that the introduction and
elimination rules (\rulest{fun} and \rulest{app}) are appropriately strong. 
This can be done by showing \emph{local soundness} and \emph{local completeness},
which correspond to $\beta$-reduction and $\eta$-expansion. Former is a special
case of subject reduction and the latter is proved by a simple derivation:
%
\begin{theorem}[Local soundness]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau$ then
$\ctyp{r}{\Gamma} \vdash \subst{e_2}{x}{e_1} : \tau$.
\end{theorem}
\begin{proof}
Special case of subject reduction (Theorem~\ref{thm:structural-reduction}).
\end{proof}
%
\begin{theorem}[Local completeness]
If $\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2$ then
$\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2$.
\end{theorem}
\begin{proof} The property is proved by the following typing derivation:
%
\begin{equation*}
\inference
  {\inference
     {\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
      \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
     {\ctyp{r \cprd (s \cvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
  {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\qedhere
\end{equation*}
\end{proof}
%
In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
guarantees that $s\cvop0 = s$. Recall that in \clflt, the typing derivation for
$\lambda x.f x$ required for local completeness was not the only possible derivation.
In the last step, it was possible to split the coeffect tag arbitrarily between the 
context and the function type.

In the \clstr~calculus, this is not, in general, the case. The $\cprd$ operator is not required 
to be associative and to have units and so a unique splitting may exist.  For example, if we
define $\cprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
$t$, there are unique $r$ and $s$ such that $t = r \cprd s$. However, if the $\cprd$ operation 
has additional properties, then there may be other possible derivation.

@@

The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
of the variable context. Thus an expression $\ctyp{r \cprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
is modelled as a function $\ctyp{r \cprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.

>>>>

\chapter{Structural coeffect language} 
\label{ch:structural}

@@

\newcommand{\csprd}[0]{ \times }
\newcommand{\csvop}[0]{ \vee }

@@

\begin{figure*}[t]
% Constants, if we wanted them, would look like this
%
% \inference[]
%   {c : \tau \in \Phi}
%   {\ctyp{1}{()} \vdash c : \tau}&
% \textnormal{\rulest{const}}
% \\[1.6em]

@@

\inference[]
  {}
  {\ctyp{0}{(v:\tau)} \vdash v : \tau }&
\textnormal{\rulest{var}}
\\[1.8em]
\inference
  {\ctyp{r \csprd s}{(\Gamma, v:\tau_1)} \vdash e : \tau_2 & v \notin \Gamma}
  {\ctyp{r}{\Gamma} \vdash \lambda v.e : \ctyp{s}{\tau_1} \rightarrow \tau_2 }&
\textnormal{\rulest{fun}}
\\[1.8em]
\inference
  {\ctyp{s}{\Gamma_1} \vdash e_1 : \ctyp{r}{\tau_1} \rightarrow \tau_2 &
   \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1 }
  {\ctyp{s \csprd (r \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash e_1~e_2 : \tau_2}&
\textnormal{\rulest{app}}
\\[1.8em]
\inference
  {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_1 : \tau_2 &
  \ctyp{t}{\Gamma_2} \vdash e_2 : \tau_1}
  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash \kvd{let}~x=e_2~\kvd{in}~e_1 : \tau_2}&
\textnormal{\rulest{let}}
\\[1.8em]
\inference
  {\ctyp{r}{\Gamma} \vdash e : \tau &
   \ctxtrans{\ctyp{r'}{\Gamma'}}{\ctyp{r}{\Gamma}}}
  {\ctyp{r'}{\Gamma'} \vdash e : \tau}&
\textnormal{\rulest{ctx}}
\end{array}
\end{equation*}

@@

\begin{equation*}
\begin{array}{rcll}
\multicolumn{3}{c}{
  \inference
    {  \ctxtrans{\ctyp{r'}{\Gamma_1'}}{\ctyp{r}{\Gamma_1}} }
    {  \ctxtrans{\ctyp{r'\csprd s}{(\Gamma_1', \Gamma_2)}}{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)}} }
}&
\textnormal{\rulest{nest}}
\\[2.0em]
  \ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)} &}{& \ctyp{s\csprd r}{(\Gamma_2, \Gamma_1)}}&
\textnormal{\rulest{exch}}
\\[0.8em]
  \ctxtrans{ \ctyp{r \csprd 1}{(\Gamma, ())} &}{&
    \ctyp{r}{\Gamma}}&
\textnormal{\rulest{empty}}
\\[0.8em]
  \ctxtrans{\ctyp{r\csprd 0}{(\Gamma, x:\tau_1)} &}{& \ctyp{r}{\Gamma}}&
\textnormal{\rulest{weak}}
\\[0.8em]
  \ctxtrans{\ctyp{r \csvop s}{(x:\tau)} &}{&
    \ctyp{r \csprd s}{(x:\tau, x:\tau)}}&
\textnormal{\rulest{contr}}
\\[0.8em]
  \ctxtrans{\ctyp{r\csprd (s\csprd t)}{(\Gamma_1, (\Gamma_2, \Gamma_3))} &}{&
    \ctyp{(r\csprd s)\csprd t}{((\Gamma_1, \Gamma_2), \Gamma_3)}}\;&
\textnormal{\rulest{assoc}}
\\[0.8em]
\multicolumn{3}{c}{
  \ctxtrans{\ctyp{r}{\Gamma}}{\ctyp{s}{\Gamma}}\quad(\textnormal{when}~s \csvop r = s)
}&
\textnormal{\rulest{sub}}
\end{array}
\end{equation*}

% We do not need this rule (can be expressed in terms of the other one)
%
% \inference
%  {  \ctxtrans{\ctyp{s}{\Gamma_2}}{\ctyp{s'}{\Gamma_2'}} }
%  {  \ctxtrans{\ctyp{r\csprd s}{(\Gamma_1, \Gamma_2)}}{\ctyp{r\csprd s'}{(\Gamma_1, \Gamma_2')}} }&
% \textnormal{\rulest{right}}

@@

As an example, consider a language that allows us to get a value of a variable (representing
some changing data-source) \ident{x} versions back using the syntax $\pastval{a}{x}$. 
To track information about individual variables, we use a product-like operation $\csprd$ on tags 
to mirrors the product structure of variables. For example:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \csprd 10}{(\ident{a}:\dnat, \ident{b}:\dnat)}
  \vdash
    \pastval{a}{5}+\pastval{b}{10}: \ident{nat}
\end{array}
\end{equation*}
%
The coeffect tag $5 \csprd 10$ corresponds to the free-variable context $\ident{a}, \ident{b}$, denoting
that we need at most 5 and 10 past values of \ident{a} and \ident{b}. If we substitute \ident{c}
for both \ident{a} and \ident{b}, we need another operation to combine multiple tags associated with 
a single variable:
%
\begin{equation*}
\begin{array}{l}
\ctyp{5 \csvop 10}{(\ident{c}:\dnat)}
  \vdash
    \pastval{c}{5}+\pastval{c}{10}: \ident{nat}
\end{array}
\end{equation*}
%
In this example, the operation $\csvop$ would be the \emph{max} function and so $5 \csvop 10 = 10$.
Before looking at the formal definition, consider the typing of let bindings:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{c} = \kvd{if}~\ident{test}()~\kvd{then}~\ident{a}~\kvd{else}~\ident{b}\\
\pastval{a}{15}+\pastval{c}{10}
\end{array}
\end{equation*}
%
The expression has free variables \ident{a} and \ident{b} (we ignore \ident{test}, which
is not a data source). It defines \ident{c}, which may be assigned either \ident{a} or \ident{b}.
The variable \ident{a} may be used directly (second line) or indirectly via \ident{c}.

The expression assigned to \ident{c} uses variables \ident{a} and \ident{b}, so its typing
context is $\ctyp{0 \csprd 0}{(\ident{a}, \ident{b})}$. The value $0$ is the unit
of $\csvop$ and it denotes empty coeffect. The typing context of the
body is $\ctyp{15 \csprd 10}{(\ident{a}, \ident{c})}$.

To combine the tags, we take the coeffect associated with \ident{c} and apply it to the tags
of the context in which \ident{c} was defined using the $\csvop$ operation. This is then combined 
with the remaining tags from the body yielding the overall context:
$\ctyp{15 \csprd (10 \csvop (0 \csprd 0))}{(\ident{a}, (\ident{a}, \ident{b}))}$. Using a simple
normalization mechanism (described later), this can be further reduced to
$\ctyp{(15 \csvop 10) \csprd 10}{(\ident{a}, \ident{b})}$. This gives us the required information --
we need at most $\emph{max}(15,10)$ past values of \ident{a} and at most $10$ past values of 
\ident{b}.

@@

\subsection{Structural coeffect tags}
In the previous section, the $\csvop$ operation behaves similarly to the flat $\vee$ operation.
However, the type system does not require some of the semilattice properties, because some uses
are replaced with the $\csprd$ operation. We do not require any properties about the $\csprd$ operation.
For example, in the previous example cannot be commutative (since a tag $15 \csprd 10$ has different 
meaning than $10 \csprd 15$). However, we relate the operations using distributivity laws to allow 
normalization that was hinted above.

\begin{definition}
A \emph{structural coeffect tag structure} $(S, \csprd, \csvop, 0, 1)$ is a tuple where 
$(S, \csvop)$ is a lattice-like structure with unit $0$. The additional structure is formed
by a binary operation $\csprd$ and element $1\in S$ such that
for all $r,s,t\in S$, the following equalities hold: 
%
\begin{subequations}
\begin{align}
\tag{associativity}
r \csvop (s \csvop t) &= (r \csvop s) \csvop t\\
\tag{commutativity}
r \csvop s &= s \csvop r\\
\tag{lower bound}
r \csvop 0 &= r\\
\tag{distributivity}
r \csvop (s \csprd t) &= (r \csvop s) \csprd (r \csvop t)\\
\tag{upper bound}
1 \csvop r &= 1
\end{align}
\end{subequations}
%
\end{definition}
%
% 'upper bound' (not sure what to call it) says that empty context cannot carry information
% looks like this is the upper bound of the bounded lattice ???
%
% commutativity is not strictly required by any law or type preservation, but
% it simplifies the presentation and looks reasonable
%
The tag $0$ represents that no coeffect is associated with a variable (i.e~when a variable is
always accessed using standard variable access). The tag $1$ is used to annotate empty 
variable context. For example, the context of an expression $\lambda x.x$ is empty and it needs
to carry an annotation. We explain exactly how this works when we introduce the type system.
The fact that $1$ is the upper bound means that combining it with other coeffect annotations
does not affect it and so empty contexts cannot carry any information.

@@

The simply typed system for \clstr~uses the same syntax of types as the system for \clflt. 
The rules are of a form $\ctyp{r}{\Gamma}\vdash e : \tau$ where $r$ is a tag provided by 
a \emph{structural coeffect tag structure} $(S,\csprd, \csvop, 0)$.

@@

As in linear and affine systems, the variable context $\Gamma$ in our system is not a 
simple set. Instead, we use the following tree-like structure (which is more similar to 
bunched types than to linear or affine systems):
%
\begin{equation*}
\Gamma ::= () \sep (x:\tau) \sep (\Gamma, \Gamma)
\end{equation*}
%
The syntax $()$ represents an empty context, so the structure defines a binary trees where
leaves are either variables or empty. Contexts such as $\ctyp{1\csprd(r \csprd 1)}{((), (x, ()))}$ 
contain unnecessary number of empty contexts $()$. However, we need to construct them
temporarily, because certain rules require splitting a context and, by our 
definition, the context $(x:\tau)$ is not splitable.

@@

The \rulest{app} rule combines coeffects $s, t, r$ associated with the function-returning
expression, argument and the function type respectively. 
The result of evaluating the argument in the context $t$ is passed to the function that
requires context $r$, so the variables used in the context $\Gamma_2$ are annotated with
the combination of coeffects $r \csvop t$. The variable context required to evaluate the 
function value is independent and so it is annotated just with coeffects $s$.
Finally, the \rulest{let} rule is derived from let binding and application, but we show
it separately to aid the understanding.

@@

\begin{lemma}[Definition of let binding]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau_2$
then $\ctyp{r}{\Gamma} \vdash \kvd{let}~x=e_1~\kvd{in}~e_2 : \tau_2$.
\end{lemma}
\begin{proof}
The premises and conclusions of a typing derivation of $(\lambda x.e_2)~e_1$
correspond with the typing rule \rulest{let}:
\begin{equation*}
\inference[]
  {\inference
    {\ctyp{r \csprd s}{(\Gamma_1, v:\tau_1)} \vdash e_2 : \tau_2 & v \notin \Gamma_1}
    {\ctyp{r}{\Gamma_1} \vdash \lambda v.e_2 : \ctyp{s}{\tau_1} \rightarrow \tau_2 } &
   \ctyp{t}{\Gamma_2} \vdash e_1 : \tau_1 }
  {\ctyp{r \csprd (s \csvop t)}{(\Gamma_1, \Gamma_2)} \vdash (\lambda v.e_2)~e_1 : \tau_2}\qedhere
\end{equation*}
\end{proof}

The term $e_2$ which is substituted in $e_1$ is checked in a different variable
and coeffect context $\ctyp{t}{\Gamma_2}$. This is common in sub-structural systems where
a variable cannot be freely used repeatedly. The context $\Gamma_2$ is used in place of 
the variable that we are substituting for. The let binding captures substitution for a 
specific variable (the context is of a form $\ctyp{r \csprd s}{\Gamma, v:\tau}$).
For a general substitution, we need to define the notion of context with a hole.

@@

\begin{equation*}
\begin{array}{rcl}
\Delta[-] &::=& \ctyp{1}{()} \\
       & \sep & \ctyp{r}{(x:\tau)}\\
       & \sep & \ctyp{-}{(-)} \\
       & \sep & \ctyp{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}\quad~~(\textnormal{where}~\ctyp{r_i}{\Gamma_i} \in \Delta[-])\\
\end{array}
\end{equation*}
%
Assuming we have a context with hole $\ctyp{r}{\Gamma} \in \Delta[-]$, the hole filling 
operation $\chole{r}{\Gamma}{r'}{\Gamma'}$ fills the hole in the variable context with 
$\Gamma'$ and the corresponding coeffect tag hole with $r'$. The operation is defined in 
Figure~\ref{fig:structural-hole-fill}. Using contexts with holes, we can now formulate the
general substitution lemma for \clstr.

@@

\begin{equation*}
\begin{array}{rcl}
                              \chole{1}{()}{r'}{\Gamma'} &=& \ctyp{1}{()}\\
                        \chole{r}{(x:\tau)}{r'}{\Gamma'} &=& \ctyp{r}{(x:\tau)}\\
                             \chole{-}{(-)}{r'}{\Gamma'} &=& \ctyp{r'}{\Gamma'}\\
\chole{r_1 \csprd r_2}{(\Gamma_1, \Gamma_2)}{r'}{\Gamma'} &=& \ctyp{r_1' \csprd r_2'}{(\Gamma_1', \Gamma_2')}\\
&\multicolumn{2}{l} {
  \textnormal{where}~~\ctyp{r_i'}{\Gamma_i'} = \chole{r_i}{\Gamma_i}{r'}{\Gamma'}
}
\end{array}
\end{equation*}
\caption{The definition of hole filling operation for $\Delta[-]$}
\label{fig:structural-hole-fill}
\vspace{-2em}
\end{figure}

\begin{lemma}[Substitution Lemma]
\label{thm:structural-subst}
If $\chole{r}{\Gamma}{R}{v:\tau'} \vdash e : \tau$ and $\ctyp{S}{\Gamma'} \vdash e' : \tau'$
then $\chole{r}{\Gamma}{R \csvop S}{\Gamma'} \vdash \subst{e}{v}{e'} : \tau$.
\end{lemma} 
\begin{proof}
Proceeds by rule induction over $\vdash$ using the properties of structural coeffect tag structure
$(S, \csvop, 0, \csprd, 1)$ (see Appendix~\ref{app:structural-substitution}).
\end{proof}

@@

\paragraph{Local soundness and completeness.}
As with the previous calculus, we want to guarantee that the introduction and
elimination rules (\rulest{fun} and \rulest{app}) are appropriately strong. 
This can be done by showing \emph{local soundness} and \emph{local completeness},
which correspond to $\beta$-reduction and $\eta$-expansion. Former is a special
case of subject reduction and the latter is proved by a simple derivation:
%
\begin{theorem}[Local soundness]
If $\ctyp{r}{\Gamma} \vdash (\lambda x.e_2)~e_1 : \tau$ then
$\ctyp{r}{\Gamma} \vdash \subst{e_2}{x}{e_1} : \tau$.
\end{theorem}
\begin{proof}
Special case of subject reduction (Theorem~\ref{thm:structural-reduction}).
\end{proof}
%
\begin{theorem}[Local completeness]
If $\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2$ then
$\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2$.
\end{theorem}
\begin{proof} The property is proved by the following typing derivation:
%
\begin{equation*}
\inference
  {\inference
     {\ctyp{r}{\Gamma} \vdash f : \ctyp{s}{\tau_1} \rightarrow \tau_2 &
      \ctyp{0}{(x:\tau_1)} \vdash x : \tau_1 }
     {\ctyp{r \csprd (s \csvop 0)}{(\Gamma, x:\tau_1)} \vdash f~x : \tau_2}}
  {\ctyp{r}{\Gamma} \vdash \lambda x.f x : \ctyp{s}{\tau_1} \rightarrow \tau_2 }
\qedhere
\end{equation*}
\end{proof}
%
In the last step, we use the \emph{lower bound} property of structural coeffect tag, which
guarantees that $s\csvop0 = s$. Recall that in \clflt, the typing derivation for
$\lambda x.f x$ required for local completeness was not the only possible derivation.
In the last step, it was possible to split the coeffect tag arbitrarily between the 
context and the function type.

In the \clstr~calculus, this is not, in general, the case. The $\csprd$ operator is not required 
to be associative and to have units and so a unique splitting may exist.  For example, if we
define $\csprd$ as the operator of a \emph{free magma}, then it is invertible and for a given
$t$, there are unique $r$ and $s$ such that $t = r \csprd s$. However, if the $\csprd$ operation 
has additional properties, then there may be other possible derivation.

@@

The semantics of structural coeffect calculus \clstr~can be defined similarly to the semantics of
\clflt. The most notable difference is that the structure of coeffect tag now mirrors the structure
of the variable context. Thus an expression $\ctyp{r \csprd s}{(\Gamma_1, \Gamma_2)} \vdash e : \tau$ 
is modelled as a function $\ctyp{r \csprd s}{(\Gamma_1 \ctimes \Gamma_2)} \rightarrow \tau$.
